---
title: "ML Prediction for Quality of Health Fitness"
author: "Ajaya Thapa"
date: "April 14, 2016"
output: 
  html_document:  
    toc: yes
---
###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The goal of this project will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Data
The training data for this project are downloaded from:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are downloaded from:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for the project come from this source: http://groupware.les.inf.puc-rio.br/har

```{r}
## Loading the required libraries
library(caret)
library(randomForest)
library(ggplot2)
```

### Loading of the Data
* Loading data.
* Remove near zero covariates and those with more than 80% missing values since these variables will not provide much power for prediction.
* Calculate correlations between each remaining feature to the response, __classe__. Use ___spearman__ rank based correlation because __classe__ is a factor.
* Plot the two features that have highest correlation with __classe__ and color with __classe__ to see if we can separate response based on these features.

```{r data, eval=TRUE}
# loading data
training <- read.csv("./data/pml-training.csv", row.names = 1)
testing <- read.csv("./data/pml-testing.csv", row.names = 1)
# remove near zero covariates
nsv <- nearZeroVar(training, saveMetrics = T)
training <- training[, !nsv$nzv]
# remove variables with more than 80% missing values
nav <- sapply(colnames(training), function(x) if(sum(is.na(training[, x])) > 0.8*nrow(training)){return(T)}else{return(F)})
training <- training[, !nav]
# calculate correlations
cor <- abs(sapply(colnames(training[, -ncol(training)]), function(x) cor(as.numeric(training[, x]), as.numeric(training$classe), method = "spearman")))
```

```{r data_plot,eval=TRUE}
# Plot Predictors
summary(cor)
plot(training[,names(which.max(cor))],training[,names(which.max(cor[-which.max(cor)]))],col=training$classe,pch=19,cex=0.1,xlab=names(which.max(cor)),ylab=names(which.max(cor[-which.max(cor)])))
```
The training set has 19622 samples and 57 potential predictors after filtering.

There doesn’t seem to be any strong predictors that correlates with __classe__ well, so linear regression model is probably not suitable in this case. Boosting and random forests algorithms may generate more robust predictions for our data.

### Boosting Model
* Fit model with boosting algorithm and 10-fold cross validation to predict classe with all other predictors.
* Plot accuracy of this model on the scale [0.9, 1].

```{r boost}
set.seed(123)
boostFit <- train(classe ~ ., method = "gbm", data = training, verbose = F, trControl = trainControl(method = "cv", number = 10))
```

```{r boost_plot}
## Plotting the Boost Fit Model
boostFit
plot(boostFit,ylim=c(0.9,1))
```

The boosting algorithm generated a good model with accuracy = 0.997.

## Random forests model
* Fit model with random forests algorithm and 10-fold cross validation to predict __classe__ with all other predictors.
* Plot accuracy of the model on the same scale as boosting model.

```{r rf,eval=F}
set.seed(123)
rfFit <- train(classe ~ ., method = "rf", data = training, importance = T, trControl = trainControl(method = "cv", number = 10))
```

```{r rf_plot}
## Plotting Random Forest Model
rfFit
plot(rfFit, ylim=c(0.9,1))
```

```{r rf_imp,echo=FALSE}
imp<-varImp(rfFit)$importance
imp$max<-apply(imp,1,max)
imp<-imp[order(imp$max,decreasing=T),]
```

The random forests algorithm generated a very accurate model with __accuracy close to 1__. Compared to boosting model, this model generally has better performance in terms of accuracy as we see from the plots.

## Final model and prediction
* Comparing model accuracy of the two models generated, random forests and boosting, random forests model has overall better accuracy. So, I’ll use this model for prediction.
* The final random forests model contains 500 trees with 40 variables tried at each split. The five most important predictors in this model are raw_timestamp_part_1, roll_belt, num_window, pitch_forearm, cvtd_timestamp30/11/2011 17:12.
* Estimated out of sample error rate for the random forests model is 0.04% as reported by the final model.
* Predict the test set and output results for automatic grader.


```{r prediction, message=FALSE}
## Final Model
rfFit$finalModel

## Prediction
(prediction<-as.character(predict(rfFit,testing)))
```

```{r results, eval=FALSE}
## Write Prediction Files
pml_write_files=function(x){
  n=length(x)
  for(i in 1:n){
    filename=paste0("/prediction/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE, row.names = FALSE, col.names=FALSE)
  }
}
pml_write_files(prediction)
```

